```{r}
library(dplyr)

library(Seurat)
library(patchwork)
library(ggplot2)
library(sf)
library(tidyr)

library(SpatEntropy)
library(spatstat)
#library(RColorBrewer)

SEED = 7
set.seed(SEED)
```
Load functions
Regional Co-occurence Entropy
```{r}
sanity_check_structs <- function(data, partition_map, partitions_sel, dist){
    #0. Sanity checks
    is.ppp(data) # returns TRUE
    is.marked(data) # returns TRUE
    if(spatstat.geom::is.ppp(data) & !spatstat.geom::is.marked(data) & length(table(data$marks$cat))==1)
        stop("Data have only 1 category. Use Batty's entropy for this use case.")
    is.list(partition_map) # returns TRUE if "areas" and FALSE if "sites"
    is.im(partition_map) # returns TRUE if "areas" and FALSE if "sites"
    if(spatstat.geom::area.owin(data$window)==1)
          stop("Nemri's entropy cannot be computed on areas of total size = 1. You should change your measurement unit so that the window size is not 1.")
}

partition_areas <- function(map, regions, map_type){
 # Get the area of each partition to table
  # if pixel image, we count the pixels in each partition
  if ( map_type == "areas") {
    T_g <- table(map$v) # Area of each partition
  }
  #if sites, we count the sites
  if ( map_type == "sites") {
    print("map")
    T_g <- table(map) # "Area" of each partition
  }
  T_g_sub <- T_g[regions] # in partitions_sel
  T_star <- min(T_g_sub) # area of smallest partition
  T_tot <- sum(T_g_sub) # total area of partitions in partitions_sel
    
  return(list(T_g, T_star, T_tot))
}

select_points_in_partitions_areas <- function(z, morpho, partitions_sel){
    # Assign variables to function-to-be
    # z has just one column
    data <- z
    partition_map <- morpho
    # 1. Assign the cells to their partition based on the factor value for the pixel where they fall
    cutdata <- cut(unmark(data), z=partition_map )
    #table(cutdata$marks) #sanity check on the cell to partition attribution - before subsetting partition of interest
    partition <- cutdata$marks
    cutdata <- unmark(cutdata)
    cutdata$marks$partition <- partition
    cutdata$marks$cat <- data$marks$cat
    cutdata$marks$p_names <- data$marks$p_names
    cutdata <- subset(cutdata, partition %in% partitions_sel) # cells in 2 and out of window are removed
    #table(cutdata$marks$marks.partition) #sanity check on the cell to partition attribution - after subsetting partition of interest

    return(cutdata)
}
select_points_in_partitions_sites <- function(z, partitions_sel){
    # z marks are 'p_names', 'cat' and 'partition
    data <- z
    cutdata <- subset(data, marks.partition %in% partitions_sel) 
    return(cutdata)
}

count_coocurrences <- function(z, dist){
  ## does not work for more than sqrt(.Machine$integer.max) = 46340 points
  ## does not consider distances between 2 points, as long as their distance is less than dist
    #library(Matrix)
    
    # Assign variables to function-to-be
    # z has just one column
    data <- z

    ct <- as.vector(data$marks$marks.cat) # vector of the categories for all the points in partitions_sel
    I = length(table(ct)) # total number of categories observed in partitions_sel
    coocs_theo <- (I*I +I)/2 # maximum number of different possible co-occurences that could be observed for m=2
    
    # 2. calculate pseudo-adjacency matrix
    eucl <- pairdist(data)  # matrix of pair-wise euclidian distances between cells
    inradius <- (eucl <= dist) # matrix of pairs of cells at a distance less than threshold = pseudo-adjacency matrix (diagonal is 1)

    #identify which pairs are in the same partition, and in which partition
    # creates n-by-n matrix where intersections are either 0 (not in the same partition) or the partition number e.g 1, 2...(ID)
    in_g <- data$marks$marks.partition
    sameg <- outer(in_g, in_g, "==") # matrix of pairs of cells that fall in the same partition

    # 3. Hadamard product of adjacency matrix and partition matrix 
    # creates n-by-n matrix where intersections are either 0 (not in the same partition) or the partition number e.g 1, 2...(ID)
    #ZLdg = eucl * inradius * sameg # turns diagonal into 0's
    #is_cooc <- (ZLdg>0) # replace all pair-wise distances with 1's
    ZLdg = inradius * sameg #ignore distances, gives diagonal of 1's
    #plot(as.im(ZLdg), clipwin=owin(xrange= c(0,100), yrange=c(0,100)))
    is_cooc <- ZLdg
    is_cooc[lower.tri(is_cooc)] <- 0 # matrix is symmetrical so co-occurences are double counted as ind1-ind2 and ind2-ind1, remove lower triangle
    is_cooc[diag(is_cooc)] <- 0 # remove diagonal to avoid counting self-links ind1-ind1
    #sum(is_cooc) # check number of found- co-occurences
    #plot(as.im(is_cooc), clipwin=owin(xrange= c(0,100), yrange=c(0,100)))
    numing <- as.numeric(in_g)
    whichg <- matrix(numing, nrow=length(numing), ncol=length(numing), byrow=TRUE)
    cooc_byg <- is_cooc * whichg # puts the partition number (ID) where previously there was 1 indicating the 2 cells are in the same partition g
    #plot(as.im(cooc_byg), clipwin=owin(xrange= c(0,100), yrange=c(0,100)))
    
    # 4. find the points in the partitions of interest and their marks
    coocs <- which(cooc_byg != 0, arr.ind = T) #gives index position as row / col of all non-zero values in is_cooc 
    ct1 <- ct[coocs[,1]] # position in row of all-vs-all matrix
    ct2 <- ct[coocs[,2]] # position in column of all-vs-all matrix
    ord_ct1 <- pmin(ct1, ct2)
    ord_ct2 <- pmax(ct1, ct2)
    ct_pairs <- paste(ord_ct1, ord_ct2, sep=" <-> ") # this is to order so that cat1-cat2 and cat2-cat1 both are counted as cat1-cat2
    #table(ct_pairs) #check
    
    # 5. make a vector of partition ids for each non zero values in coocs_byg
    cp <- as.vector(data$marks$marks.partition)
    g <- cp[coocs[,1]] # position in row of all-vs-all matrix # (sum(g == cp[coocs[,2]]) == length(ct_pairs)) # Returns TRUE, no need to order
    ct_pairs_byg <- data.frame(ct_pairs, ord_ct1, ord_ct2, g)
    #print(ct_pairs_byg)
    
    count_p_g <- ct_pairs_byg %>%
      group_by(ct_pairs,ord_ct1, ord_ct2,g) %>% summarize(n = n()) 
    
    return(count_p_g)
}

prep_b4_calc_H <- function(df, rep_sim, min_coocs){
  
  df["n"] <- df[rep_sim]
  df[rep_sim] <- NULL
  
  #remove pairs that have no occurences across all g's
  #start by identifying pairs with no occurences for safekeeping
  df_miss <- df %>% filter(sum_n == 0)
  nb_pairs_miss <- length(unique(df_miss$ct_pairs))
  #print(nb_pairs_miss)
  df <- df %>% filter(sum_n >0)
  #print(df)
  nb_pairs <- length(unique(df$ct_pairs))
  #print(nb_pairs)
  
  # replace NA's by 0 to prepare to add 1
  df$n[is.na(df$n)] <- 0
 
  #identify edge case of pairs not occuring in at least one partition
  undist <- df %>% filter( n == 0) 
  uneven_pairs_ls <- unique(undist$ct_pairs)
  #print(undist)
  # add 1 to occurence to each partition where there the pair does not occur in at least one partition
  df$n_mod <- df$n +  (df$ct_pairs %in% uneven_pairs_ls)
  rm(undist, uneven_pairs_ls)
  
  # recalculate total occurences per pair
  ncoocs_by_ctpairs <-  df[,c("ct_pairs", "n_mod")] %>% # Count the total number of cells from each co-occurence (across all partitions)
      group_by(ct_pairs) %>% summarise(sum_n_mod= sum(n_mod))
  df <- inner_join(df, ncoocs_by_ctpairs, by="ct_pairs")
  
  # remove pairs where total occurences less than set minimum
  df <-  df %>% filter( sum_n_mod > min_coocs) 
  
  df["n"] <- df["n_mod"]
  df["sum_n"] <- df["sum_n_mod"]
  df["n_mod"] <- NULL
  df["sum_n_mod"] <- NULL
  
  return(df)
}

counts_to_H <- function(df, T_all){

    T_star <- T_all[[2]]
    T_tot <- T_all[[3]]
  
    df$p = df$n / df$sum_n
    df$plogtg_p = df$p * log(df$Tg / df$p)
    
    dec_H_r <- df %>%
      group_by(ct_pairs) %>% summarise(Hr = sum(plogtg_p)) 
    
   # dec_H$n <- ncoocs_by_ctpairs$Cells_in_ct_pairs
    dec_H_r$Hr_rel <- (dec_H_r$Hr - log(T_star)) / (log(T_tot) - log(T_star))
    
    #plot(dec_H_r$Hr_rel)
    #summary(dec_H_r$Hr_rel)
    
    return(list(df, dec_H_r))
}

```

Create work directory
```{r}
workdir <- "~/mouse_brain_alzheimer"
if (!dir.exists(workdir)){
  dir.create(workdir)
}else{
  print("dir exists")
}
setwd(workdir)
```
Download data from https://cf.10xgenomics.com/samples/xenium/1.4.0/Xenium_V1_FFPE_TgCRND8_17_9_months/Xenium_V1_FFPE_TgCRND8_17_9_months_outs.zip
Unzip to workdir
```{r}
data.dir <- paste(workdir, "Xenium_V1_FFPE_TgCRND8_17_9_months_outs", sep = "/")
xenium.obj <- LoadXenium(data.dir, fov = "fov", assay = "Xenium")
# remove cells with 0 counts
xenium.obj <- subset(xenium.obj, subset = nCount_Xenium > 0)
#ImageDimPlot(xenium.obj, fov = "fov", molecules = c("B2m", "Pecam1"), nmols = 20000, flip_xy = FALSE)
#VlnPlot(xenium.obj, features = c("nFeature_Xenium", "nCount_Xenium"), ncol = 2, pt.size = 0)

xenium.obj <- SCTransform(xenium.obj, assay = "Xenium")
xenium.obj <- RunPCA(xenium.obj, npcs = 30, features = rownames(xenium.obj))
xenium.obj <- RunUMAP(xenium.obj, dims = 1:30)
xenium.obj <- FindNeighbors(xenium.obj, reduction = "pca", dims = 1:30)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)

Idents(xenium.obj) <- xenium.obj$seurat_clusters
xenium.obj.markers <- FindAllMarkers(xenium.obj, only.pos = FALSE)

xenium.obj.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() -> top20
```


```{r}
#Fig S1a
pdf("~/mouse_brain_alzheimer/figs1a.pdf",     fonts = c("Helvetica"),    # File name
    width = 14 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to #inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
   paper = "A4")          # Paper size
p <- ImageDimPlot(xenium.obj, group.by = "ident", cols = "alphabet", size =  0.5, flip_xy = FALSE, dark.background = FALSE) + #, border.color = "black", border.size=0.2) + 
  labs(fill = "Cell types") +
  theme(
    panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_text(family = "Helvetica", size = 7), 
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white", colour = "black", linewidth = 0)) 
p
dev.off()
#ImageDimPlot(xenium.obj, size = 2, flip_xy = FALSE) + scale_fill_brewer(palette="Paired") +theme(panel.background = element_rect(fill="black"))

#Fig S1b
pdf("~/mouse_brain_alzheimer/figs1b.pdf",         # File name
    width = 12 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
p <- DimPlot(xenium.obj,  group.by = "ident",  cols = "alphabet", raster=TRUE) +#, label=TRUE, repel = TRUE ) +
  theme(
    plot.title = element_blank(),
    #legend.position = "none",
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"),
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white"),#, colour = "white", linewidth = 0),
    label.text = element_text(family = "Helvetica", size = 6),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
LabelClusters(p, id="ident", repel = TRUE, family = "Helvetica", size = 2, color = "black")
p
dev.off()

library(patchwork)

# identify cells corresponding to astrocytes
#Fig S1c
pdf("~/mouse_brain_alzheimer/figs1c.pdf",         # File name
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
ps<- list()
for (gene in c("Gja1", "Ntsr2", "Aqp4", "Acsbg1", "Slc39a12")) {
p <- FeaturePlot(object = xenium.obj, features = gene, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
ps[[gene]]<- p
}
layout1<-"
AB
CD
E#
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()



# identify the cluster corresponding to microglia
#Fig S1d
pdf("~/mouse_brain_alzheimer/figs1d.pdf",         # File name
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
ps<- list()
for (gene in c("C1qa", "C1qb", "C1qc", "Ctss", "Csf1r")) {
p <- FeaturePlot(object = xenium.obj, features = gene, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
ps[[gene]]<- p
}
layout1<-"
AB
CD
E#
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()



SaveSeuratRds(xenium.obj, file = "mouse_brain_after_clustering.rds")
```
```{r}
xenium.obj <- LoadSeuratRds("mouse_brain_after_clustering.rds")
```

SELECT RELEVANT CELL TYPES
```{r}
# Subset cells 
selection <- c("1", "4")
xenium.obj.sel <- subset(xenium.obj, subset = seurat_clusters %in% selection)
xenium.obj.sel$seurat_clusters_rd1 <- xenium.obj.sel$seurat_clusters

xenium.obj.sel <- SCTransform(xenium.obj.sel, assay = "Xenium")
xenium.obj.sel <- RunPCA(xenium.obj.sel, npcs = 30, features = rownames(xenium.obj.sel))
xenium.obj.sel <- RunUMAP(xenium.obj.sel, dims = 1:30, repulsion.strength = 30)
xenium.obj.sel <- FindNeighbors(xenium.obj.sel, reduction = "pca", dims = 1:30)

DimPlot(xenium.obj.sel, label=TRUE, repel = TRUE )

combined <- xenium.obj.sel
## Select a range of resolutions
resolution.range <- seq(from = 0, to = 1, by = 0.1)
## Find clusters using a range of resolutions
combined <- FindClusters(object = combined, resolution = resolution.range)
library(clustree)
clustree(combined)
rm(combined)


xenium.obj.sel <- FindClusters(xenium.obj.sel, resolution = 1) # ######### 0.7) ########################
xenium.obj.sel.markers <- FindAllMarkers(xenium.obj.sel, only.pos = FALSE) 

xenium.obj.sel.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() -> top20_subcluster_1_4

manual_v1 <- c(
  "Astrocytes reactive A1 (Fiber tracts)",# 0: Gsn, Vim, Cd44, C3
  "Astrocytes (Cortex; Dkk3)",# 1 : # Increases neuronal damage
  "Microglia DAMs (Protective, CDCc11+OPN-)",# 2 : Trem2, Tyrobp
  "Astrocytes (Hypothalamus; Calb2)",# 3:  ##################
  "Astrocytes (Thalamus; Slc39a12)",# 4: ################################
  "Microglia Homeostatic (Cortex)",# 5: P2ry12, Tmem119 ############
  "Microglia Homeostatic (Cortex)",# 6: P2ry12, Tmem119 ###########
  "Astrocytes (Hypothalamus; Calb2)",# 7 : ############
  "Neurons",# 8: Shisa6, Cabp7
  "Microglia Homeostatic (Thalamus)",# 9: Cx3cr1, P2ry12, Tmem119
  "Microglia DAMs (Pathogenic, CDCc11+OPN+)",# 10: Spp1, Itgax, Gpnmb
  "Astrocytes reactive group 2",# 11: Gfap, Vim, Serpina3n, Cd44
  "Microglia Homeostatic (Hippocampus)",# 12: Cx3cr1, Tmem119 #############
  "Astrocytes (Striatum; Penk)",# 13: Penk,
  "Astrocytes (Cortex; Dkk3)",# 14 : # Increases neuronal damage ############
  "Astrocytes (Cxcl10)",# 15: Cxcl10, Vim
  "Neurons",# 16: Cbln1, Cbln4 # with adipocytes Slc17a6
  "Neurons" # 17: Shisa6, Cabp7
)
xenium.obj.sel$celltypes_manual_v1 <- manual_v1[as.factor(xenium.obj.sel$seurat_clusters)]

ImageDimPlot(xenium.obj.sel, group.by = "celltypes_manual_v1", cols = "Paired", size = 1.5, flip_xy = FALSE)
DimPlot(xenium.obj.sel, size = 2, group.by = "celltypes_manual_v1", label=TRUE, repel = TRUE )

#keep only astrocytes and microglia - exclude Neurons clusters
immune <- subset(xenium.obj.sel, subset = seurat_clusters %in% c("8", "16", "17") , invert =TRUE)

immune_order1 <- c(
  "Astrocytes (Cortex; Dkk3)",
  "Astrocytes (Cxcl10)",
  "Astrocytes (Hypothalamus; Calb2)",
  "Astrocytes (Striatum; Penk)",
  "Astrocytes (Thalamus; Slc39a12)",
  "Astrocytes reactive A1 (Fiber tracts)",
  "Astrocytes reactive group 2",
  "Microglia DAMs (Pathogenic, CDCc11+OPN+)",
  "Microglia DAMs (Protective, CDCc11+OPN-)",
  "Microglia Homeostatic (Cortex)",
  "Microglia Homeostatic (Hippocampus)",
  "Microglia Homeostatic (Thalamus)"
)
Idents(immune) <- immune$celltypes_manual_v1
levels(immune) <- immune_order1

SaveSeuratRds(immune, file = "mouse_brain_after_2nd_clustering.rds")
```

```{r}
immune <- LoadSeuratRds("mouse_brain_after_2nd_clustering.rds")
```


```{r}
colors = brewer.pal(n=length(levels(as.factor(immune$celltypes_manual_v1))), name = "Paired")

#Fig 1c
df.ct <- as.data.frame(table(immune$celltypes_manual_v1))
pdf("~/mouse_brain_alzheimer/fig1c.pdf",         # File name
    width = 6 / 2.54 , height = 7 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
p <- ggplot(df.ct, aes(x=forcats::fct_rev(Var1), y=Freq)) + 
  geom_bar(stat="identity", fill = colors) + 
  scale_x_discrete(labels = function(x)  stringr::str_wrap(x, width = 15)) +   coord_flip() +
    labs(x="Cell counts", y="Cell types") + 
    theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black", hjust=0.5, lineheight=0.7),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
        panel.grid.major = element_line(color="grey", linewidth = 0.5),
        panel.grid.minor = element_line(color="grey", linewidth = 0.25),
    )
p
dev.off()

#Fig S2a
pdf("~/mouse_brain_alzheimer/figs2a.pdf",         # File name
    width = 18 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
p <- ImageDimPlot(immune, group.by = "ident", cols = colors, size = 0.5, flip_xy = FALSE, dark.background = FALSE, border.color = "black", border.size=0.2) + 
  labs(fill = "Cell types") +
  theme(
    panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_text(family = "Sans", size = 7), 
    legend.text = element_text(family = "Sans", size = 6),
    legend.key = element_rect(fill = "white", colour = "black", linewidth = 0)) 
    #panel.border =  element_rect(color='black'))
p
dev.off()

#Fig S2b
pdf("~/mouse_brain_alzheimer/figs2b.pdf",         # File name
    width = 15 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
p <- DimPlot(immune,  group.by = "ident",  pt.size = 2, cols = colors, raster=TRUE) +#, label=TRUE, repel = TRUE ) + +
  theme(
    plot.title = element_blank(),
    #legend.position = "none",
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"),
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white"),#, colour = "white", linewidth = 0),
    label.text = element_text(family = "Helvetica", size = 6),
    )
#LabelClusters(p, id="ident", repel = TRUE, family = "Helvetica", size = 2, color = "black")
#p$layers[[1]]$aes_params$stroke <- 0
#p$layers[[1]]$aes_params$size <- 4
p
dev.off()

#Fig S2c
pdf("~/mouse_brain_alzheimer/figs2c.pdf",         # File name
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
ps<- list()
for (gene in c("Vim", "Gsn", "Serpina3n", "Dkk3", "Calb2", "Penk")) {
p <- FeaturePlot(object = immune, features = gene, pt.size=5, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
#p$layers[[1]]$aes_params$stroke <- 0
#p$layers[[1]]$aes_params$size <- 0.1
ps[[gene]]<- p
}
layout1<-"
AB
CD
EF
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()

#Fig S2d
pdf("~/mouse_brain_alzheimer/figs2d.pdf",         # File name
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
ps<- list()
for (gene in c("C1qa", "C1qb", "C1qc", "Itgax", "Trem2", "Spp1")) {
p <- FeaturePlot(object = immune, features = gene, pt.size=5, raster = TRUE) + 
    theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    #legend.key.height= unit(0.1, 'cm'),
    #legend.key.width = unit(0.2, "cm"),
    #legend.position = c (0.1,0.2),
    legend.position = "none",
    )
#p$layers[[1]]$aes_params$stroke <- 0
#p$layers[[1]]$aes_params$size <-1
ps[[gene]]<- p
}
layout1<-"
AB
CD
EF
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()


```

```{r}
cat_factor <- "celltypes_manual_v1" 
ma <- immune
#rm(xenium.obj,xenium.obj.sel)
```
Convert cells to marked point pattern
```{r}
centroids <- ma[["fov"]]$centroids
ma.maxx <- max(centroids$x)
ma.maxy <-max(centroids$y)
#the observation is controlled by the cell locations, not the IF image
w <- owin(c(0,ma.maxx), c(0,ma.maxy), unitname=c("um", "ums"))
z <- ppp(centroids$x, centroids$y, window = w) #, marks = colnames(ma))
marks(z) <- data.frame(p_names = rownames(ma[[]]), cat = pull(ma[[]], cat_factor))
```
The original IF is first aligned in Xenium Explorer, from which a 3x3 alignment matrix is exported.
This is in case the morphology and spatial transcriptomics object are not aligned, i.e same origin (0,0),
scale and orientation.
As per the Xenium alignment tutorial, the transformation matrix has translation, scaling, rotation, and reflection.
The transformation value units are in pixels, and the image coordinates and its annotations are in microns. The (0,0) position is in the top left corner of the image.
We use 'affine' from spatstat which takes a 2x2 matrix for rotation/reflection, and 2-values vector for translation. We then scale accounting for difference in original pixel sizes between the 2 images used for alignment.

Original IF image 
width : 15527 px (5046.08 um)
height : 19540 px (6350.30 um)
IFpxwidth = 0.3250
IFpxheight = 0.3250

Other ome files (same origin/scale as Xenium object)
31348 px (6661.45 um)
24864 px (5283.60 um)
OMEpx width: 0.2125 um
OMEpx height : 0.2125 um

Save alignment file as "Xenium_V1_FFPE_TgCRND8_17_9_months_if_image_alignment_files/matrix.csv"
```{r}
alignment_matrix_file <- paste(workdir, "Xenium_V1_FFPE_TgCRND8_17_9_months_if_image_alignment_files/matrix.csv", sep="/")
#alignment_mat <- c(-0.005620739234351479,1.5290391356596393,2166.8103540491984, 
#                   -1.5290391356596393,	-0.005620739234351479,	22201.00617577853,
#                   0,	0,	1)
#align <- matrix(alignment_mat, ncol=3, byrow=TRUE)
#
IFpxwidth = 0.3250
IFpxheight = 0.3250
OMEpxwidth =  0.2125
OMEpxheight = 0.2125

align <- as.matrix(read.csv(alignment_matrix_file, header=F))
aff <- align[c(1,2), c(1,2)]
transx <- align[,3][1] * IFpxwidth
transy <-  align[,3][2] * IFpxheight

pt_geojson <- "~/mouse_brain_alzheimer/amyloid_expansion_30.geojson"
# crs to specify geographic coordinates are in planar system not spheric (default)
annotations_sf <- st_read(pt_geojson, crs = NA)
geo <- st_geometry(annotations_sf) 
windows <- lapply(geo, as.owin)
df <- st_drop_geometry(annotations_sf)
te <- tess(tiles=windows)
marks(te) <- df
te
#plot(te, do.col = FALSE, do.labels = TRUE, main = "")
#plot(te, values = marks(te)$name, main = "")

reduction_factor <- 16
xmin <- 0 #xmin <- te$window$xrange[1]
xmax <- 15527 #xmax <- te$window$xrange[2]
ymin <- 0 #ymin <- te$window$yrange[2]
ymax <- 19540 #ymax <- te$window$yrange[2]

morpho <- as.im(te, dimyx=c(xmax/reduction_factor, ymax/reduction_factor))
#for some reason, we have to rescale twice. tried multiple combination incl. scaling by 1/OMEpxwidth to no avail
plot(morpho, col=c('cyan', 'lightpink', 'white'), col.alpha=0.5)
morpho <- rescale.im(morpho, 1/IFpxwidth, "ums")
morpho <- affine(morpho, mat= aff, vec=c(transx, transy ))
morpho <- rescale.im(morpho,IFpxwidth/OMEpxwidth, "ums")

#check that the two plots align
#plot(morpho, col=c('grey90', 'gray20', 'white'), col.alpha=0.5)
#plot(z, pch=19, cex=0.1, alpha = 1, use.marks=FALSE, add=TRUE )
#check that the two plots align
palette = "Paired" 
colors = brewer.pal(n=length(levels(as.factor(z$marks$cat))), name = palette)
# Fig. 1D
pdf("~/mouse_brain_alzheimer/fig1d.pdf",         # File name
    width = 24 / 2.54 , height = 20 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
# run all 3 lines to create overlayed plots
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat", cols = colors  , legend=TRUE, leg.side="left", leg.args=list(cex=2, nsymbols = length(colors)), show.window=FALSE) #
plot(morpho, col=c('white', 'gray70', 'black'), col.alpha=0.5, ribbon=TRUE,  add=TRUE)
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat"  , cols = colors, add=TRUE, show.window=FALSE)
dev.off()
```
```{r}
#Parameters setting
map_type <- "areas"
dist = 30
if ( map_type == "sites"){
  dist = 0
}
partitions_sel = c(1,3) # cells in 2 and out of window are removed
min_coocs= 50
iter_n <- 1000

# initialize results dataframe
uni_cat <- unique(z$marks$cat)
r_g <- expand.grid(uni_cat, uni_cat, as.character(partitions_sel))
ord_ct1 <- pmin(as.character(r_g$Var1), as.character(r_g$Var2))
ord_ct2 <- pmax(as.character(r_g$Var1), as.character(r_g$Var2))
ct_pairs <- paste(ord_ct1, ord_ct2, sep=" <-> ")
r_g_theo <- data.frame(ct_pairs, ord_ct1, ord_ct2, r_g$Var3)
r_g_theo <- r_g_theo %>% distinct()
colnames(r_g_theo) <- c('ct_pairs', 'ord_ct1', 'ord_ct2', 'g')

# collect partition area
T_all <- partition_areas(morpho, partitions_sel, map_type)
T_g <- T_all[[1]]
rownames(T_g) <- as.character(rownames(T_g))
r_g_theo$Tg <- T_g[as.character(r_g_theo$g)]

# 1. get results for the actual data
z_real <- z
#sanity_check_structs(z_real, morpho, partitions_sel, dist)
if (map_type == "areas") {
  z_real <- select_points_in_partitions_areas(z_real, morpho, partitions_sel)
}
if (map_type == "sites") {
  z_real <- select_points_in_partitions_sites(z_real, partitions_sel)
}
countpg <- count_coocurrences(z_real, dist)
countpg$ct_pairs <- NULL
names(countpg)[names(countpg) == 'n'] <- 'real.counts'
r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))


# 2. add the random permutations
for (iter in 1:iter_n) {
  print(iter)
  z_shuf <- z
  z_shuf$marks$cat <- sample(z_shuf$marks$cat)
  # check where cells fall
  #table(z_shuf$marks$marks.partition)/z_shuf$n
  
  #sanity_check_structs(z_shuf, morpho, partitions_sel, dist)
  if (map_type == "areas") {
    z_shuf <- select_points_in_partitions_areas(z_shuf, morpho, partitions_sel)
  }
  if (map_type == "sites") {
    z_shuf <- select_points_in_partitions_sites(z_shuf, partitions_sel)
  }
  countpg <- count_coocurrences(z_shuf, dist)
  countpg$ct_pairs <- NULL
  names(countpg)[names(countpg) == 'n'] <- paste('rand', iter, sep="_")
  r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))

}

#3. Save raw results from real data and permutation
write.csv(r_g_theo, "~/mouse_brain_alzheimer/coocs_1000_permutations_v2.csv")
```


```{r}
#r_g_theo <- read.csv("~/mouse_brain_alzheimer/coocs_1000_permutations_v2.csv") 


#4. Convert to summary results
reps <- c("real.counts", paste(sep="", "rand_", 1:iter_n)) 
cols = c("ct_pairs", "ord_ct1", "ord_ct2", "g", "Tg")

df <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df) <- c('Hr_rel', 'rep')

for (rep_sim in reps) {
  print(rep_sim)
  r_allg <- r_g_theo[,c(cols, rep_sim)]
  ncoocs_by_ctpairs <- r_g_theo[,c(cols, rep_sim)] %>% # Count the total number of cells from each co-occurence (across all partitions)
      replace(is.na(.), 0) %>% 
      group_by(ct_pairs) %>% summarise(sum_n= sum(get(rep_sim)))
  r_allg <- inner_join(r_allg, ncoocs_by_ctpairs, by="ct_pairs")
  r_allg <- arrange(r_allg, ct_pairs, g)
  r_all_g_fltr <-  prep_b4_calc_H(r_allg, rep_sim, min_coocs)
  dec_H_rep <- counts_to_H(r_all_g_fltr, T_all)
  r_all_g_fltr = dec_H_rep[[1]]
  dec_H_r = dec_H_rep[[2]]
  #store results from this rep in dataframe
  df <- rbind(df, data.frame(Hr_rel = dec_H_r$Hr_rel, rep = rep_sim))
  if (rep_sim == "real.counts"){ 
    df.real.counts <- dec_H_rep
  }
}

real.counts.ordered <- arrange(df.real.counts[[2]],Hr_rel)
write.csv(real.counts.ordered, "~/mouse_brain_alzheimer/real.counts.Hr_rel_v2.csv")
```

```{r}
T_star <- T_all[[2]]
T_tot <- T_all[[3]]

realcnts <- df.real.counts[[1]] %>% 
  group_by(ct_pairs) %>% 
  summarise(sum_n = sum(n), RCE = sum(plogtg_p), g1 = ifelse(g == 1, n, 0), g2 = ifelse(g == 2, n, 0), g3 = ifelse(g == 3, n, 0) ) %>%  
  arrange(,RCE)
realcnts <- realcnts %>%
  group_by(ct_pairs) %>%
  summarise(RCE = max(RCE), relRCE= (max(RCE) - log(T_star)) / (log(T_tot) - log(T_star)),
            n = max(sum_n),  g1 = sum(g1), g2 = sum(g2), g3 = sum(g3) ) %>% 
  arrange(,RCE)
#Fig. 2B
print(n=150, realcnts)
```



```{r}
#real.counts.ordered <- read.csv("~/mouse_brain_alzheimer/real.counts.Hr_rel_v2.csv") 
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

df_by_rep <- df %>% group_by(rep) %>% summarise(mean_Hr_rel= mean(Hr_rel))
df_by_rep$test <- with(df_by_rep, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
permutations <- df_by_rep[df_by_rep$test == "permutation test",]
#df$test <- with(df, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
summary(permutations$mean_Hr_rel)
var(permutations$mean_Hr_rel)

#Fig 2a
pdf("~/mouse_brain_alzheimer/fig2a.pdf",         # File name
    width = 4 / 2.54 , height = 5 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
p <- ggplot(df_by_rep, aes(x=test, y=mean_Hr_rel)) + 
  geom_boxplot(size = 0.5, colour="black", fill="white", linewidth = 0.5, outlier.size = 0.1) +
  geom_jitter(size=0.1, alpha = 0.3) +
  #theme(legend.position="none") +
  #labs(title="Alzheimer Example") +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\ncell types","permuted\ncell types")) +
  #coord_cartesian(ylim = c(0.91, 1)) +  
  scale_y_continuous(name =  "Relative RCE", breaks=seq(0.91,0.99,0.01), limits=c(0.915, 0.999), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 



#Fig. 2b
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

#Fig. S3a
min(df[df$test == "permutation test",]$Hr_rel)
#df$rep <- factor(df$rep , levels=reps)
df$test <- with(df, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
pdf("~/mouse_brain_alzheimer/figs3a.pdf",         # File name
    width = 4 / 2.54 , height = 5 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",          # Background color
    colormodel = "cmyk",    # Color model (cmyk is required for most publications)
    paper = "A4")          # Paper size
library(ggforce)
p <- ggplot(df, aes(x=test, y=Hr_rel, fill=test)) +   geom_violin() +
  #geom_point(position = position_jitter(seed = 1, width = 0.2)) + 
  geom_point(size = 0.5, fill="black", alpha=1) + 
  geom_sina(alpha=0.01) +
  #geom_jitter(position = position_jitter(seed = 1, width = 0.2)) +
  theme(legend.position="none") +
  #labs(title="Bird Example") +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\ncell types","permuted\ncell types")) +
  #coord_cartesian(ylim = c(0.91, 1)) +  
  scale_y_continuous(name =  "Relative RCE", breaks=seq(0.6,0.99,0.1), limits=c(0.5, 1), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 




```
