---
title: "alzheimer_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{alzheimer_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(ggforce)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(Seurat)
library(scales)
library(sf)
library(spatstat)

library(RC.entropy)

SEED = 7
set.seed(SEED)
```

Create work directory
```{r}
workdir <- "~/mouse_brain_alzheimer_v2"

if (!dir.exists(workdir)){
  dir.create(workdir)
}else{
  print("dir exists")
}
#setwd(workdir) # this breaks knitr during devtools::check()
```

//// SEURAT WORKFLOW 1/2

This next block is a classical single-cell workflow for 10X Genomics Xenium data using Seurat.
It serves to cluster all cells present in the sample.
The resulting mouse_brain_after_clustering.rds file is saved as the workflow is time-consuming, enabling the user to skip this 1st step after a 1st run.

Download data from https://cf.10xgenomics.com/samples/xenium/1.4.0/Xenium_V1_FFPE_TgCRND8_17_9_months/Xenium_V1_FFPE_TgCRND8_17_9_months_outs.zip
Unzip to workdir
```{r}
data.dir <- paste(workdir, "Xenium_V1_FFPE_TgCRND8_17_9_months_outs", sep = "/")
xenium.obj <- LoadXenium(data.dir, fov = "fov", assay = "Xenium")

xenium.obj <- subset(xenium.obj, subset = nCount_Xenium > 0) # remove cells with 0 counts
xenium.obj <- SCTransform(xenium.obj, assay = "Xenium")
xenium.obj <- RunPCA(xenium.obj, npcs = 30, features = rownames(xenium.obj))
xenium.obj <- RunUMAP(xenium.obj, dims = 1:30)
xenium.obj <- FindNeighbors(xenium.obj, reduction = "pca", dims = 1:30)
xenium.obj <- FindClusters(xenium.obj, resolution = 0.3)
Idents(xenium.obj) <- xenium.obj$seurat_clusters
xenium.obj.markers <- FindAllMarkers(xenium.obj, only.pos = FALSE)
xenium.obj.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() -> top20

SaveSeuratRds(xenium.obj, file = "mouse_brain_after_clustering.rds")
```

START HERE by uncommenting the 1st line if you performed the Seurat workflow 1/2 previously and wish to skip it

```{r}
#xenium.obj <- LoadSeuratRds("mouse_brain_after_clustering.rds")
```

GENERATE FIGURE S1 from the manuscript

```{r}
#Fig S1a
pdf(paste(workdir, "figs1a.pdf", sep = "/"),     fonts = c("Helvetica"), 
    width = 14 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to #inches
    bg = "white",         
    colormodel = "cmyk",    
   paper = "A4") 
p <- ImageDimPlot(xenium.obj, group.by = "ident", cols = "alphabet", size =  0.5, flip_xy = FALSE, dark.background = FALSE) +
  labs(fill = "Cell types") +
  theme(
    panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_text(family = "Helvetica", size = 7), 
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white", colour = "black", linewidth = 0)) 
p
dev.off()

#Fig S1b
pdf(paste(workdir, "figs1b.pdf", sep = "/"),
    width = 12 / 2.54 , height = 8 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",
    colormodel = "cmyk",
    paper = "A4") 
p <- DimPlot(xenium.obj,  group.by = "ident",  cols = "alphabet", raster=TRUE) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"),
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white"),
    label.text = element_text(family = "Helvetica", size = 6),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
LabelClusters(p, id="ident", repel = TRUE, family = "Helvetica", size = 2, color = "black")
p
dev.off()

# identify cells corresponding to astrocytes
#Fig S1c
pdf(paste(workdir, "figs1c.pdf", sep = "/"),
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",
    colormodel = "cmyk",
    paper = "A4") 
ps<- list()
for (gene in c("Gja1", "Ntsr2", "Aqp4", "Acsbg1", "Slc39a12")) {
p <- FeaturePlot(object = xenium.obj, features = gene, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
ps[[gene]]<- p
}
layout1<-"
AB
CD
E#
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()

# identify the cluster corresponding to microglia
#Fig S1d
pdf(paste(workdir, "figs1d.pdf", sep = "/"), 
    width = 8 / 2.54 , height = 10 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",
    colormodel = "cmyk",
    paper = "A4") 
ps<- list()
for (gene in c("C1qa", "C1qb", "C1qc", "Ctss", "Csf1r")) {
p <- FeaturePlot(object = xenium.obj, features = gene, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
p$layers[[1]]$aes_params$stroke <- 0
p$layers[[1]]$aes_params$size <- 0.1
ps[[gene]]<- p
}
layout1<-"
AB
CD
E#
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()
```

//// SEURAT WORKFLOW 2/2
 
The next blocks are a classical single-cell workflow for 10X Genomics Xenium data using Seurat.
It serves to select cells, namely astrocytes and microglia, identified in the previous workflow. to perform the 2d round of clustering on selected cells,  identified 
The resulting mouse_brain_after_2nd_clustering.rds file is saved as the workflow is time-consuming, enabling the user to skip this 2nd step after a 1st run.

```{r}
selection <- c("1", "4") # Subset cells 
xenium.obj.sel <- subset(xenium.obj, subset = seurat_clusters %in% selection)
xenium.obj.sel$seurat_clusters_rd1 <- xenium.obj.sel$seurat_clusters
xenium.obj.sel <- SCTransform(xenium.obj.sel, assay = "Xenium")
xenium.obj.sel <- RunPCA(xenium.obj.sel, npcs = 30, features = rownames(xenium.obj.sel))
xenium.obj.sel <- RunUMAP(xenium.obj.sel, dims = 1:30, repulsion.strength = 30)
xenium.obj.sel <- FindNeighbors(xenium.obj.sel, reduction = "pca", dims = 1:30)
DimPlot(xenium.obj.sel, label=TRUE, repel = TRUE )

xenium.obj.sel <- FindClusters(xenium.obj.sel, resolution = 1)
xenium.obj.sel.markers <- FindAllMarkers(xenium.obj.sel, only.pos = FALSE) 
xenium.obj.sel.markers %>%
    group_by(cluster) %>%
    dplyr::filter(avg_log2FC > 1) %>%
    slice_head(n = 20) %>%
    ungroup() -> top20_subcluster_1_4

manual_v1 <- c(
  "Astrocytes reactive A1 (Fiber tracts)",# 0: Gsn, Vim, Cd44, C3
  "Astrocytes (Cortex; Dkk3)",# 1 : # Increases neuronal damage
  "Microglia DAMs (Protective, CDCc11+OPN-)",# 2 : Trem2, Tyrobp
  "Astrocytes (Hypothalamus; Calb2)",# 3:  
  "Astrocytes (Thalamus; Slc39a12)",# 4: 
  "Microglia Homeostatic (Cortex)",# 5: P2ry12, Tmem119 
  "Microglia Homeostatic (Cortex)",# 6: P2ry12, Tmem119
  "Astrocytes (Hypothalamus; Calb2)",# 7 : 
  "Neurons",# 8: Shisa6, Cabp7
  "Microglia Homeostatic (Thalamus)",# 9: Cx3cr1, P2ry12, Tmem119
  "Microglia DAMs (Pathogenic, CDCc11+OPN+)",# 10: Spp1, Itgax, Gpnmb
  "Astrocytes reactive group 2",# 11: Gfap, Vim, Serpina3n, Cd44
  "Microglia Homeostatic (Hippocampus)",# 12: Cx3cr1, Tmem119 
  "Astrocytes (Striatum; Penk)",# 13: Penk,
  "Astrocytes (Cortex; Dkk3)",# 14 : # Increases neuronal damage 
  "Astrocytes (Cxcl10)",# 15: Cxcl10, Vim
  "Neurons",# 16: Cbln1, Cbln4 # with adipocytes Slc17a6
  "Neurons" # 17: Shisa6, Cabp7
)
xenium.obj.sel$celltypes_manual_v1 <- manual_v1[as.factor(xenium.obj.sel$seurat_clusters)]

ImageDimPlot(xenium.obj.sel, group.by = "celltypes_manual_v1", cols = "Paired", size = 1.5, flip_xy = FALSE)
DimPlot(xenium.obj.sel, size = 2, group.by = "celltypes_manual_v1", label=TRUE, repel = TRUE )

#keep only astrocytes and microglia - exclude Neurons clusters
immune <- subset(xenium.obj.sel, subset = seurat_clusters %in% c("8", "16", "17") , invert =TRUE)

immune_order1 <- c(
  "Astrocytes (Cortex; Dkk3)",
  "Astrocytes (Cxcl10)",
  "Astrocytes (Hypothalamus; Calb2)",
  "Astrocytes (Striatum; Penk)",
  "Astrocytes (Thalamus; Slc39a12)",
  "Astrocytes reactive A1 (Fiber tracts)",
  "Astrocytes reactive group 2",
  "Microglia DAMs (Pathogenic, CDCc11+OPN+)",
  "Microglia DAMs (Protective, CDCc11+OPN-)",
  "Microglia Homeostatic (Cortex)",
  "Microglia Homeostatic (Hippocampus)",
  "Microglia Homeostatic (Thalamus)"
)
Idents(immune) <- immune$celltypes_manual_v1
levels(immune) <- immune_order1

SaveSeuratRds(immune, file = "mouse_brain_after_2nd_clustering.rds")
```

START HERE by uncommenting the next line if you performed the Seurat workflow 2/2 previously and wish to skip it

```{r}
immune <- LoadSeuratRds(paste(workdir, "mouse_brain_after_2nd_clustering.rds", sep="/"))
```

GENERATE FIGURES 1C, S2 from the manuscript
```{r}
colors = brewer.pal(n=length(levels(as.factor(immune$celltypes_manual_v1))), name = "Paired")

#Fig 2c
df.ct <- as.data.frame(table(immune$celltypes_manual_v1))
pdf(paste(workdir, "fig2c.pdf", sep = "/"),
    width = 6 / 2.54 , height = 7 / 2.54,
    bg = "white", 
    colormodel = "cmyk",
    paper = "A4")
p <- ggplot(df.ct, aes(x=forcats::fct_rev(Var1), y=Freq)) + 
  geom_bar(stat="identity", fill = colors) + 
  scale_x_discrete(labels = function(x)  stringr::str_wrap(x, width = 15)) +   coord_flip() +
    labs(x="Cell counts", y="Cell types") + 
    theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black", hjust=0.5, lineheight=0.7),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
        panel.grid.major = element_line(color="grey", linewidth = 0.5),
        panel.grid.minor = element_line(color="grey", linewidth = 0.25),
    )
p
dev.off()

#Fig S2a
pdf(paste(workdir, "figs2a.pdf", sep = "/"),
    width = 18 / 2.54 , height = 8 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
p <- ImageDimPlot(immune, group.by = "ident", cols = colors, size = 0.5, flip_xy = FALSE, dark.background = FALSE, border.color = "black", border.size=0.2) + 
  labs(fill = "Cell types") +
  theme(
    panel.background = element_rect(fill="white", color="black", linewidth = 0.5),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.title = element_text(family = "Sans", size = 7), 
    legend.text = element_text(family = "Sans", size = 6),
    legend.key = element_rect(fill = "white", colour = "black", linewidth = 0)) 
p
dev.off()

#Fig S2b
pdf(paste(workdir, "figs2b.pdf", sep = "/"),
    width = 15 / 2.54 , height = 8 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
p <- DimPlot(immune,  group.by = "ident",  pt.size = 2, cols = colors, raster=TRUE) +
  theme(
    plot.title = element_blank(),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"),
    legend.text = element_text(family = "Helvetica", size = 6),
    legend.key = element_rect(fill = "white"),#, colour = "white", linewidth = 0),
    label.text = element_text(family = "Helvetica", size = 6),
    )
p
dev.off()

#Fig S2c
pdf(paste(workdir, "figs2c.pdf", sep = "/"),
    width = 8 / 2.54 , height = 10 / 2.54,
    bg = "white", 
    colormodel = "cmyk",
    paper = "A4")
ps<- list()
for (gene in c("Vim", "Gsn", "Serpina3n", "Dkk3", "Calb2", "Penk")) {
p <- FeaturePlot(object = immune, features = gene, pt.size=5, raster = TRUE) + 
  theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.key.height= unit(0.1, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    legend.position = c (0.1,0.2),
    )
ps[[gene]]<- p
}
layout1<-"
AB
CD
EF
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()

#Fig S2d
pdf(paste(workdir, "figs2d.pdf", sep = "/"),
    width = 8 / 2.54 , height = 10 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
ps<- list()
for (gene in c("C1qa", "C1qb", "C1qc", "Itgax", "Trem2", "Spp1")) {
p <- FeaturePlot(object = immune, features = gene, pt.size=5, raster = TRUE) + 
    theme(
    plot.title = element_text(family = "Helvetica", size = 6, face = "bold", color = "black"),
    axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
    axis.title.x = element_text(family = "Helvetica", size =5, color = "black"),
    axis.title.y = element_text(family = "Helvetica", size = 5, color = "black"),
    legend.text=element_text(family = "Helvetica", size = 5, color = "black"),
    axis.line = element_line(color = "black",linewidth = 0.25),
    axis.ticks = element_line(color = "black",linewidth = 0.25),
    legend.position = "none",
    )
ps[[gene]]<- p
}
layout1<-"
AB
CD
EF
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()

```

/// FORMAT CONVERSION 1/2 FROM SPATIAL OMICS DATA TO SPATIAL POINT PATTERN 

Generate spatial point pattern (z). It uses cells centroid coordinates from .rds object and cell types (celltypes_manual_v1) annotated previously.

```{r}

cat_factor <- "celltypes_manual_v1" 
ma <- immune

centroids <- ma[["fov"]]$centroids
ma.maxx <- max(centroids$x)
ma.maxy <-max(centroids$y)
#the observation window is defined by the minimum and maximum coordinates of cell locations, not the IF image
w <- owin(c(0,ma.maxx), c(0,ma.maxy), unitname=c("um", "ums"))
z <- ppp(centroids$x, centroids$y, window = w) 
marks(z) <- data.frame(p_names = rownames(ma[[]]), cat = pull(ma[[]], cat_factor))
```

/// FORMAT CONVERSION 2/2 FROM MORPHOLOGY IMAGE TO MARKED PIXEL IMAGE (PARTITIONS) 

Generate regional map (morpho).

The original IF is first aligned in Xenium Explorer, from which a 3x3 alignment matrix is exported.
Save alignment file as alignment_matrix.csv (after exporting in folder "Xenium_V1_FFPE_TgCRND8_17_9_months_if_image_alignment_files/matrix.csv" from Xenium Explorer)

This is in case the morphology and spatial transcriptomics object are not aligned, i.e they don't share the same origin (0,0), scale and orientation.
As per the Xenium alignment tutorial, the transformation matrix has translation, scaling, rotation, and reflection.
The transformation value units are in pixels, and the image coordinates and its annotations are in microns. The (0,0) position is in the top left corner of the image.
We use 'affine' from spatstat which takes a 2x2 matrix for rotation/reflection, and 2-values vector for translation. We then scale accounting for difference in original pixel sizes between the 2 images used for alignment.

Original IF image 
width : 15527 px (5046.08 um)
height : 19540 px (6350.30 um)
IFpxwidth = 0.3250
IFpxheight = 0.3250

Other ome files (same origin/scale as Xenium object)
31348 px (6661.45 um)
24864 px (5283.60 um)
OMEpx width: 0.2125 um
OMEpx height : 0.2125 um


```{r}
alignment_matrix_file <- paste(workdir, "alignment_matrix.csv", sep="/")

IFpxwidth = 0.3250
IFpxheight = 0.3250
OMEpxwidth =  0.2125
OMEpxheight = 0.2125

align <- as.matrix(read.csv(alignment_matrix_file, header=F))
aff <- align[c(1,2), c(1,2)]
transx <- align[,3][1] * IFpxwidth
transy <-  align[,3][2] * IFpxheight

pt_geojson <- paste(workdir, "amyloid_expansion_30.geojson", sep="/")
annotations_sf <- st_read(pt_geojson, crs = NA) # crs to specify geographic coordinates are in planar system not spheric (default)
geo <- st_geometry(annotations_sf) 
windows <- lapply(geo, as.owin)
df <- st_drop_geometry(annotations_sf)
te <- tess(tiles=windows)
marks(te) <- df

reduction_factor <- 16
xmin <- 0 #xmin <- te$window$xrange[1]
xmax <- 15527 #xmax <- te$window$xrange[2]
ymin <- 0 #ymin <- te$window$yrange[2]
ymax <- 19540 #ymax <- te$window$yrange[2]

morpho <- as.im(te, dimyx=c(xmax/reduction_factor, ymax/reduction_factor))
#for some reason, we have to rescale twice. tried multiple combination incl. scaling by 1/OMEpxwidth to no avail
plot(morpho, col=c('cyan', 'lightpink', 'white'), col.alpha=0.5)
# select regions to keep from the visualisation generated above (partitions_sel = c(1,3))
morpho <- rescale.im(morpho, 1/IFpxwidth, "ums")
morpho <- affine(morpho, mat= aff, vec=c(transx, transy ))
morpho <- rescale.im(morpho,IFpxwidth/OMEpxwidth, "ums")

#check that the two plots align
palette = "Paired" 
colors = brewer.pal(n=length(levels(as.factor(z$marks$cat))), name = palette)
# Fig. 1b
pdf(paste(workdir, "fig1b.pdf", sep = "/"),       
    width = 24 / 2.54 , height = 20 / 2.54, 
    bg = "white", 
    colormodel = "cmyk", 
    paper = "A4") 
# run all 3 lines to create overlayed plots
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat", cols = colors  , legend=TRUE, leg.side="left", leg.args=list(cex=2, nsymbols = length(colors)), show.window=FALSE) #
plot(morpho, col=c('white', 'gray70', 'black'), col.alpha=0.5, ribbon=TRUE,  add=TRUE)
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat"  , cols = colors, add=TRUE, show.window=FALSE)
dev.off()
```
/// REGIONAL CO-OCCURENCE ENTROPY workflow starts here

```{r}
#Parameters setting
map_type <- "areas" # alternative is: "sites"
dist = 30 # this is the parameter "d"
if ( map_type == "sites"){
  dist = 0
  morpho <- NULL
}
partitions_sel = c(1,3) # see block above where morpho is plotted to select regions of interest
min_coocs= 50
iter_n <- 1000 # number of random permutations of labels to run 

# initialize results dataframe
uni_cat <- unique(z$marks$cat)
r_g <- expand.grid(uni_cat, uni_cat, as.character(partitions_sel))
ord_ct1 <- pmin(as.character(r_g$Var1), as.character(r_g$Var2))
ord_ct2 <- pmax(as.character(r_g$Var1), as.character(r_g$Var2))
ct_pairs <- paste(ord_ct1, ord_ct2, sep=" <-> ")
r_g_theo <- data.frame(ct_pairs, ord_ct1, ord_ct2, r_g$Var3)
r_g_theo <- r_g_theo %>% distinct()
colnames(r_g_theo) <- c('ct_pairs', 'ord_ct1', 'ord_ct2', 'g')

# collect partition area
T_all <- partition_areas(morpho, partitions_sel, map_type)
T_g <- T_all[[1]]
rownames(T_g) <- as.character(rownames(T_g))
r_g_theo$Tg <- T_g[as.character(r_g_theo$g)]

# 1. get results for the actual data - you may want to stop there if you are not interested in testing permutations
z_real <- z
#sanity_check_structs(z_real, morpho, partitions_sel, dist) #perform sanity checks, TO DO properly
if (map_type == "areas") {
  z_real <- select_points_in_partitions_areas(z_real, morpho, partitions_sel)
}
if (map_type == "sites") {
  z_real <- select_points_in_partitions_sites(z_real, partitions_sel)
}
countpg <- count_coocurrences(z_real, dist)
countpg$ct_pairs <- NULL
names(countpg)[names(countpg) == 'n'] <- 'real.counts'
r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))


# 2. add the random permutations
for (iter in 1:iter_n) {
  print(iter)
  z_shuf <- z
  z_shuf$marks$cat <- sample(z_shuf$marks$cat)
  # check where cells fall
  #table(z_shuf$marks$marks.partition)/z_shuf$n
  
  #sanity_check_structs(z_shuf, morpho, partitions_sel, dist)
  if (map_type == "areas") {
    z_shuf <- select_points_in_partitions_areas(z_shuf, morpho, partitions_sel)
  }
  if (map_type == "sites") {
    z_shuf <- select_points_in_partitions_sites(z_shuf, partitions_sel)
  }
  countpg <- count_coocurrences(z_shuf, dist)
  countpg$ct_pairs <- NULL
  names(countpg)[names(countpg) == 'n'] <- paste('rand', iter, sep="_")
  r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))

}

#3. Save raw results from real data and permutation
perm <- paste("coocs",iter_n, "permutations.csv", sep="_")
write.csv(r_g_theo, paste(workdir, perm, sep="/"))
```

Checkpoint after iterations

```{r}
# Uncomment this line if you have already the time-consuming step if iterations of random permutations
#r_g_theo <- read.csv( paste(workdir, perm, sep="/")) 

#4. Convert to summary results
reps <- c("real.counts", paste(sep="", "rand_", 1:iter_n)) 
cols = c("ct_pairs", "ord_ct1", "ord_ct2", "g", "Tg")

df <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df) <- c('Hr_rel', 'rep')

for (rep_sim in reps) {
  print(rep_sim)
  r_allg <- r_g_theo[,c(cols, rep_sim)]
  ncoocs_by_ctpairs <- r_g_theo[,c(cols, rep_sim)] %>% # Count the total number of cells from each co-occurence (across all partitions)
      replace(is.na(.), 0) %>% 
      group_by(ct_pairs) %>% summarise(sum_n= sum(get(rep_sim)))
  r_allg <- inner_join(r_allg, ncoocs_by_ctpairs, by="ct_pairs")
  r_allg <- arrange(r_allg, ct_pairs, g)
  r_all_g_fltr <-  prep_b4_calc_H(r_allg, rep_sim, min_coocs)
  dec_H_rep <- counts_to_H(r_all_g_fltr, T_all)
  r_all_g_fltr = dec_H_rep[[1]]
  dec_H_r = dec_H_rep[[2]]
  #store results from this rep in dataframe
  df <- rbind(df, data.frame(Hr_rel = dec_H_r$Hr_rel, rep = rep_sim))
  if (rep_sim == "real.counts"){ 
    df.real.counts <- dec_H_rep
  }
}

real.counts.ordered <- arrange(df.real.counts[[2]],Hr_rel)
write.csv(real.counts.ordered,  paste(workdir, "real.counts.Hr_rel.csv", sep="/"))
```

Decomposed RCE's
Sort by lowest (most signal) to highest (lowest signal=random) and display just the top pairs
```{r}
T_star <- T_all[[2]]
T_tot <- T_all[[3]]

realcnts <- df.real.counts[[1]] %>% 
  group_by(ct_pairs) %>% 
  summarise(sum_n = sum(n), RCE = sum(plogtg_p), g1 = ifelse(g == 1, n, 0), g2 = ifelse(g == 2, n, 0), g3 = ifelse(g == 3, n, 0) ) %>%  
  arrange(,RCE)
realcnts <- realcnts %>%
  group_by(ct_pairs) %>%
  summarise(RCE = max(RCE), relRCE= (max(RCE) - log(T_star)) / (log(T_tot) - log(T_star)),
            n = max(sum_n),  g1 = sum(g1), g2 = sum(g2), g3 = sum(g3) ) %>% 
  arrange(,RCE)
#Fig. 2B
print(n=150, realcnts)
```

Summary statistics of permutations
Plotting of permutations against real data averaged per iteration (fig 2a) and with each decomposed RCE (fig s3a)
Decomposed RCE's of pairs (fig 2b)

```{r}
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

df_by_rep <- df %>% group_by(rep) %>% summarise(mean_Hr_rel= mean(Hr_rel))
df_by_rep$test <- with(df_by_rep, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
permutations <- df_by_rep[df_by_rep$test == "permutation test",]

summary(permutations$mean_Hr_rel)
var(permutations$mean_Hr_rel)

#Fig 2d
pdf(paste(workdir, "fig2d.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
p <- ggplot(df_by_rep, aes(x=test, y=mean_Hr_rel)) + 
  geom_boxplot(size = 0.5, colour="black", fill="white", linewidth = 0.5, outlier.size = 0.1) +
  geom_jitter(size=0.1, alpha = 0.3) +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\ncell types","permuted\ncell types")) +
  scale_y_continuous(name =  "Relative RCE", breaks=seq(0.88,0.99,0.02), limits=c(0.885, 0.999), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 


#Fig. 2b
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

#Fig. 2e
min(df[df$test == "permutation test",]$Hr_rel)
df$test <- with(df, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
pdf(paste(workdir, "fig2e.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54,
    bg = "white",
    colormodel = "cmyk", 
    paper = "A4") 
p <- ggplot(df, aes(x=test, y=Hr_rel, fill=test)) +   geom_violin() +
  geom_point(size = 0.5, fill="black", alpha=1) + 
  geom_sina(alpha=0.01) +
  theme(legend.position="none") +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\ncell types","permuted\ncell types")) +
  scale_y_continuous(name = "Decomposed relative RCE", breaks=seq(0.6,0.99,0.1), limits=c(0.5, 1), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 



```

