---
title: "birds_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{birds_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(ggforce)
library(ggplot2)
library(grid)
library(gridExtra)
library(patchwork)
library(RColorBrewer)
library(scales)
library(scatterpie)
library(sf)
library(spatstat)
library(tidyr)

library(RC.entropy)

library(spAbundance)

SEED = 7
set.seed(SEED)
```

Create work directory
```{r}
workdir <- "~/birds_neon_v2"

if (!dir.exists(workdir)){
  dir.create(workdir)
}else{
  print("dir exists")
}
#setwd(workdir) # this breaks knitr during devtools::check()
```

90 sites with $coords$easting and $coords$northing (longitude, latitude)
within a site all individuals have the same coordinates, we set map_type = "sites"  causing dist = 0
```{r}


data(neonDWP)
sp.names <- dimnames(neonDWP$y)[[1]]
abundance_ac4dists <- rowSums(neonDWP$y, dims = 2) # the original matrix has a 3rd dimension with 4-dimensional arrays for the distance breaks, collapse
sum(abundance_ac4dists) #753
tr_ab <- t(abundance_ac4dists)

easting <- neonDWP$coords[, "easting"]
northing <- neonDWP$coords[, "northing"]
forest <- neonDWP$covs$forest
grass <- neonDWP$covs$grass
tr_ab <- cbind(tr_ab, easting, northing, forest, grass)
tr_ab <- as.data.frame(tr_ab)
tr_ab$other_cover <- 1- tr_ab$forest - tr_ab$gras

split_f = 1.75
#environments must be stored as numbers
envs <- matrix(c(1, 2, 3), ncol=3, byrow=TRUE)
colnames(envs) <-  c("forest", "grass", "mixed")
tr_ab$environment <- ifelse( tr_ab$forest > split_f * tr_ab$grass , envs[1], ifelse(tr_ab$grass > split_f * tr_ab$forest , envs[2], envs[3]))
tr_ab$environment_n <- ifelse( tr_ab$forest > split_f * tr_ab$grass , "forest_dominant",  ifelse(tr_ab$grass > split_f * tr_ab$forest , "grass_dominant", "mixed_dominance"))
#tr_ab$forest_dominant <- tr_ab$forest > 1.75 * tr_ab$grass

pforest <- ggplot(neonDWP$covs, aes(x=wind, y=forest)) + geom_violin()  + geom_jitter(shape=16, position=position_jitter(0.2))
pgrass  <- ggplot(neonDWP$covs, aes(x=wind, y=grass)) + geom_violin() + geom_jitter(shape=16, position=position_jitter(0.2))
pforest + pgrass
plot(neonDWP$covs$forest,neonDWP$covs$grass )
min(neonDWP$coords[,"easting"])

# Fig. 3ab and s4
pdf(paste(workdir, "fig3a_s4.pdf", sep = "/"),
    width = 20 / 2.54 , height = 20 / 2.54, # Width and height in centimeters converted to inches
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
ps<- list()
ps[[1]]  <- ggplot() + geom_scatterpie(aes(x=easting, y=northing), data=tr_ab, cols = sp.names, 
                                       pie_scale = 1.25, linewidth=0.1, ) + coord_equal() +
  labs(title= "Species composition of sites", x = "Easting (meters)", y = "Northing (meters)") +
  labs(fill = "Species") +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        plot.title = element_blank(),
        legend.title = element_text(family = "Helvetica", size = 7),
        legend.text = element_text(family = "Helvetica", size = 5),
        legend.key = element_rect(fill = "white"),
        legend.key.height= unit(0.2, 'cm'),
        legend.key.width = unit(0.2, "cm"),
        panel.background = element_rect(fill = "white"),
        panel.border = element_rect(fill = NA, colour="black", linewidth = 0.5),
        panel.grid.major = element_line(colour="lightgrey", size=0.5),
        panel.grid.minor = element_line(colour="lightgrey", size=0.25),
        ) +
  theme(aspect.ratio=1.5)

ps[[2]]  <- ggplot() + geom_scatterpie(aes(x=easting, y=northing), data=tr_ab, cols = c("forest", "grass", "other_cover"),
                                       pie_scale = 1.25, linewidth=0.1, ) + coord_equal()  +
  scale_fill_manual(values=c("chartreuse2", "gold", "grey80")) +
  labs(x = "Easting (meters)", y = "Northing (meters)")+
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        #panel.grid.minor = element_blank()
        legend.title = element_text(family = "Helvetica", size = 7),
        legend.text = element_text(family = "Helvetica", size = 5),
    legend.key = element_rect(fill = "white"),#, colour = "white", linewidth = 0),
    legend.key.height= unit(0.2, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(fill = NA, colour="black", linewidth = 0.5),
    panel.grid.major = element_line(colour="lightgrey", size=0.5),
    panel.grid.minor = element_line(colour="lightgrey", size=0.25),
        ) +
  theme(aspect.ratio=1.5)

ps[[3]] <- ggplot(tr_ab, aes(x=easting, y=northing, col = environment_n),
                  ) + coord_equal() +
  geom_point(size=2.5) +
  labs(x = "Easting (meters)", y = "Northing (meters)") +
    theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        #panel.grid.minor = element_blank()
        legend.title = element_text(family = "Helvetica", size = 7),
        legend.text = element_text(family = "Helvetica", size = 5),
    legend.key = element_rect(fill = "white"),#, colour = "white", linewidth = 0),
    legend.key.height= unit(0.2, 'cm'),
    legend.key.width = unit(0.2, "cm"),
    panel.background = element_rect(fill = "white"),
    panel.border = element_rect(fill = NA, colour="black", linewidth = 0.5),
    panel.grid.major = element_line(colour="lightgrey", size=0.5),
    panel.grid.minor = element_line(colour="lightgrey", size=0.25),
        ) +
  theme(aspect.ratio=1.5)

layout1<-"
A#
BC
"
wrap_plots(ps ,guides = 'collect', design = layout1)
dev.off()

```


Expand counts per site into individual points and convert to spatial point pattern
Gives warning for duplicated points, which is expected behaviour for 'sites' data
```{r}
options(digits=16)
#expand summary table to prepare data for spatial point pattern
df <- data.frame(matrix(ncol = 6, nrow = 0))
colnames(df) <- c('bird', 'species', 'site', 'easting', 'northing', 'environment')

for( i in rownames(tr_ab) ) {
  for ( j in sp.names ) {
    count_sp_site = tr_ab[i,j]
    if ( count_sp_site > 0 )
      for (k in 1:count_sp_site){
        indiv <- paste(j, i, k, sep="_")
        row_add = c(indiv, j, i, tr_ab[i, 'easting'], tr_ab[i, 'northing'], tr_ab[i, "environment"])
        df[nrow(df) + 1,] = row_add
      }
  }
}
df$easting <- as.numeric(df$easting)
df$northing <- as.numeric(df$northing)
df

#Convert cells to marked point pattern
win.minx <- min(df$easting)
win.maxx <- max(df$easting)
win.miny <- min(df$northing)
win.maxy <- max(df$northing)

w <- owin(c(win.minx,win.maxx), c(win.miny,win.maxy), unitname=c("meter", "meters"))
z <- ppp(df$easting, df$northing, window = w)
marks_df = df[,c('bird', 'species', 'environment')]
colnames(marks_df) <- c('marks.p_names', 'marks.cat', 'marks.partition')
marks(z) <- marks_df
```

/// REGIONAL CO-OCCURENCE ENTROPY workflow starts here

```{r}
# Parameters setting
map_type <- "sites"
if ( map_type == "sites"){
  dist = 0
  morpho <- NULL
}
min_coocs= 3
iter_n <- 1000
partitions_sel <- c(1, 2, 3)


# initialize results dataframe 
uni_cat <- unique(z$marks$marks.cat)
r_g <- expand.grid(uni_cat, uni_cat, as.character(partitions_sel))
ord_ct1 <- pmin(as.character(r_g$Var1), as.character(r_g$Var2))
ord_ct2 <- pmax(as.character(r_g$Var1), as.character(r_g$Var2))
ct_pairs <- paste(ord_ct1, ord_ct2, sep=" <-> ")
r_g_theo <- data.frame(ct_pairs, ord_ct1, ord_ct2, r_g$Var3)
r_g_theo <- r_g_theo %>% distinct()
colnames(r_g_theo) <- c('ct_pairs', 'ord_ct1', 'ord_ct2', 'g')

#####################################
#in this case, the areas do not matter, as the birds are recorded for the whole site, with the center of the observation circle (radius 250m)
#being recorded as coordinates for all the birds (overlapping points)
#the area can then just be the number of sites of a given environment-type
# collect partition area
T_all <- partition_areas(tr_ab$environment, partitions_sel, map_type)
T_g <- T_all[[1]]
rownames(T_g) <- as.character(rownames(T_g))
r_g_theo$Tg <- T_g[as.character(r_g_theo$g)]

# 1. get counts for the actual data
z_real <- z

#sanity_check_structs(z_real, morpho, partitions_sel, dist) # TO DO add extra checks
if (map_type == "areas") {
  z_real <- select_points_in_partitions_areas(z_real, morpho, partitions_sel)
}
if (map_type == "sites") {
  z_real <- select_points_in_partitions_sites(z_real, partitions_sel)
}
countpg <- count_coocurrences(z_real, dist)
countpg$ct_pairs <- NULL
names(countpg)[names(countpg) == 'n'] <- 'real.counts'
r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))


# 2. add the random permutations
for (iter in 1:iter_n) {
  print(iter)
  z_shuf <- z
  z_shuf$marks$marks.cat <- sample(z_shuf$marks$marks.cat)
  # check where birds fall
  table(z_shuf$marks$marks.partition)/z_shuf$n
  
  #sanity_check_structs(z_shuf, morpho, partitions_sel, dist) # TO DO add extra checks
  if (map_type == "areas") {
    z_shuf <- select_points_in_partitions_areas(z_shuf, morpho, partitions_sel)
  }
  if (map_type == "sites") {
    z_shuf <- select_points_in_partitions_sites(z_shuf, partitions_sel)
  }
  countpg <- count_coocurrences(z_shuf,  dist)
  countpg$ct_pairs <- NULL
  names(countpg)[names(countpg) == 'n'] <- paste('rand', iter, sep="_")
  r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))

}

#3. Save raw results from real data and permutation
perm <- paste("coocs",iter_n, "permutations.csv", sep="_")
write.csv(r_g_theo, paste(workdir, perm, sep="/"))

```

Checkpoint after iterations

```{r}
# Uncomment this line if you have already done the time-consuming step of iterations of random permutations
#r_g_theo <- read.csv( paste(workdir, perm, sep="/"))

#4. Convert to summary results
reps <- c("real.counts", paste(sep="", "rand_", 1:iter_n)) 
cols = c("ct_pairs", "ord_ct1", "ord_ct2", "g", "Tg")

df <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df) <- c('Hr_rel', 'rep')

for (rep_sim in reps) {
  print(rep_sim)
  r_allg <- r_g_theo[,c(cols, rep_sim)]
  ncoocs_by_ctpairs <- r_g_theo[,c(cols, rep_sim)] %>% # Count the total number of cells from each co-occurence (across all partitions)
      replace(is.na(.), 0) %>% 
      group_by(ct_pairs) %>% summarise(sum_n= sum(get(rep_sim)))
  r_allg <- inner_join(r_allg, ncoocs_by_ctpairs, by="ct_pairs")
  r_allg <- arrange(r_allg, ct_pairs, g)
  r_all_g_fltr <-  prep_b4_calc_H(r_allg, rep_sim, min_coocs)
  dec_H_rep <- counts_to_H(r_all_g_fltr, T_all)
  r_all_g_fltr = dec_H_rep[[1]]
  dec_H_r = dec_H_rep[[2]]
  #store results from this rep in dataframe
  df <- rbind(df, data.frame(Hr_rel = dec_H_r$Hr_rel, rep = rep_sim))
  if (rep_sim == "real.counts"){ 
    df.real.counts <- dec_H_rep
  }
}

real.counts.ordered <- arrange(df.real.counts[[2]],Hr_rel)
write.csv(real.counts.ordered,  paste(workdir, "real.counts.Hr_rel.csv", sep="/"))

```

Decomposed RCE's
Sort by lowest (most signal) to highest (lowest signal=random) and display just the top pairs

```{r}
T_star <- T_all[[2]]
T_tot <- T_all[[3]]

realcnts <- df.real.counts[[1]]  %>% 
  group_by(ct_pairs) %>% 
  summarise(sum_n = sum(n), RCE = sum(plogtg_p), g1 = ifelse(g == 1, n, 0), g2 = ifelse(g == 2, n, 0), g3 = ifelse(g == 3, n, 0) ) %>%  
  arrange(,RCE)
realcnts <- realcnts %>%
  group_by(ct_pairs) %>%
  summarise(RCE = max(RCE), relRCE= (max(RCE) - log(T_star)) / (log(T_tot) - log(T_star)),
            n = max(sum_n),  g1 = sum(g1), g2 = sum(g2), g3 = sum(g3) ) %>% 
  arrange(,RCE)
#fig 2d (and table s2)
print(n=150, realcnts)


```
Summary statistics of permutations
Plotting of permutations against real data averaged per iteration (fig 2c) and with each decomposed RCE (fig s3b)
Decomposed RCE's of pairs (fig ??)

```{r}
# Uncomment this line if you have already the time-consuming step if iterations of random permutations
#r_g_theo <- read.csv( paste(workdir, perm, sep="/")) 
#real.counts.ordered <- read.csv(paste(workdir, "real.counts.Hr_rel.csv", sep="/"))

print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

df_by_rep <- df %>% group_by(rep) %>% summarise(mean_Hr_rel= mean(Hr_rel))
df_by_rep$test <- with(df_by_rep, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
permutations <- df_by_rep[df_by_rep$test == "permutation test",]

summary(permutations$mean_Hr_rel)
var(permutations$mean_Hr_rel)

#Fig. 3d
pdf(paste(workdir, "fig3d.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54, 
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
p <- ggplot(df_by_rep, aes(x=test, y=mean_Hr_rel)) + 
  geom_boxplot(size = 0.5, colour="black", fill="white", linewidth = 0.5, outlier.size = 0.1) +
  geom_jitter(size=0.1, alpha = 0.3) +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\nspecies","permuted\nspecies")) +
  scale_y_continuous(name = "Relative RCE", breaks=seq(0.88,0.99,0.02), limits=c(0.885, 0.999), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 

# Fig. 3e
pdf(paste(workdir, "fig3e.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4") 
library(ggforce)
p <- ggplot(df, aes(x=test, y=Hr_rel, fill=test)) +   geom_violin() +
  geom_point(size = 0.1, fill="black", alpha=0.1) + 
  geom_sina(alpha=0.01) +
  theme(legend.position="none") +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\nspecies","permuted\nspecies")) +
  scale_y_continuous(name =   "Decomposed relative RCE", breaks=seq(0.6,0.99,0.1), limits=c(0.5, 1), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 

```
