---
title: "village_example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{village_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(sf)
library(spatstat)

library(RC.entropy)

SEED = 7
set.seed(SEED)
```

Create work directory
```{r}
workdir <- "~/dennery_drone"

if (!dir.exists(workdir)){
  dir.create(workdir)
}else{
  print("dir exists")
}
#setwd(workdir) # this breaks knitr during devtools::check()
```

/// FORMAT CONVERSION 1/2 FROM CENTROIDS GEOSON TO SPATIAL POINT PATTERN 

Generate spatial point pattern (z) and partition map (morpho). 

```{r}
# 1. load partitions and convert to pixel image
g_geojson <- paste(workdir, "area_blocks_epsg3857.geojson", sep="/")
g_sf <- st_read(g_geojson, crs = NA) # crs to specify geographic coordinates are in planar system not spheric (default)
p <- g_sf[2]
plot(p)
win <- st_bbox(p)
#one polygon is empty, remove it
pne <- p %>% filter(!st_is_empty(.))
geo <- st_geometry(pne) 
windows <- lapply(geo, as.owin)
df <- st_drop_geometry(pne)
te <- tess(tiles=windows)
marks(te) <- df
reduction_factor <- 2
xmin <- te$window$xrange[1]
xmax <- te$window$xrange[2]
ymin <- te$window$yrange[1]
ymax <- te$window$yrange[2]
morpho <- as.im(te, dimyx=c((xmax-xmin)/reduction_factor, (ymax-ymin)/reduction_factor))

# 2. load buildings and convert to spatial point pattern
cat_factor <- "ROOF_MATERIAL"
pt_geojson <- paste(workdir, "dennery_centroids.geojson", sep="/")
annotations_sf <- st_read(pt_geojson, crs = NA) # crs to specify geographic coordinates are in planar system not spheric (default)
z <- as.ppp(annotations_sf)
z$window <- te$window
marks(z) <- data.frame(p_names = rownames(z$marks), cat = pull(z$marks, cat_factor))

# 3. check that the two plots align
z_palette = "Set1" 
z_colors = brewer.pal(n=length(levels(as.factor(z$marks$cat))), name = z_palette)
m_palette = "Pastel1"
m_colors = brewer.pal(n=length(levels(as.factor(morpho$v))), name = m_palette)
# Fig. X1
pdf(paste(workdir, "figX1.pdf", sep = "/"),       
    width = 24 / 2.54 , height = 20 / 2.54, 
    bg = "white", 
    colormodel = "cmyk", 
    paper = "A4") 
# run all 3 lines to create overlayed plots
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat", cols = z_colors  , legend=TRUE, leg.side="left", leg.args=list(cex=2, nsymbols = length(z_colors)), show.window=FALSE) #
plot(morpho, col=m_colors, col.alpha=0.5, ribbon=TRUE,  add=TRUE)
plot(z, pch=19, cex=0.05, alpha = 1, use.marks=TRUE, which.marks = "cat"  , cols = z_colors, add=TRUE, show.window=FALSE)
dev.off() 
 

```

```{r}
# cat 1
conditions <- c("BLUE_TARP", "CONCRETE_CEMENT", "HEALTHY_METAL")
replacement_values <- c("intact", "intact", "intact", "intact")
z$marks$cat2 <- replace(z$marks$cat, z$marks$cat %in% conditions, replacement_values)
# cat 2
conditions <- c("INCOMPLETE", "IRREGULAR_METAL")
replacement_values <- c("damaged", "damaged")
z$marks$cat <- replace(z$marks$cat2, z$marks$cat2 %in% conditions, replacement_values)
z$marks$cat2 <- NULL

#check frequencies of the different categories (not split by partitions)
table(z$marks$cat)
```
/// REGIONAL CO-OCCURENCE ENTROPY workflow starts here
```{r}
#Parameters setting
map_type <- "areas" # alternative is: "sites"
dist = 15 # this is the parameter "d"
if ( map_type == "sites"){
  dist = 0
  morpho <- NULL
}
partitions_sel = c(1,2,3,4,5,6) # see block above where morpho is plotted to select regions of interest
min_coocs= 10
iter_n <- 1000 # number of random permutations of labels to run 

# initialize results dataframe
uni_cat <- unique(z$marks$cat)
r_g <- expand.grid(uni_cat, uni_cat, as.character(partitions_sel))
ord_ct1 <- pmin(as.character(r_g$Var1), as.character(r_g$Var2))
ord_ct2 <- pmax(as.character(r_g$Var1), as.character(r_g$Var2))
ct_pairs <- paste(ord_ct1, ord_ct2, sep=" <-> ")
r_g_theo <- data.frame(ct_pairs, ord_ct1, ord_ct2, r_g$Var3)
r_g_theo <- r_g_theo %>% distinct()
colnames(r_g_theo) <- c('ct_pairs', 'ord_ct1', 'ord_ct2', 'g')

# collect partition area information
T_all <- partition_areas(morpho, partitions_sel, map_type)
T_g <- T_all[[1]]
rownames(T_g) <- as.character(rownames(T_g))
r_g_theo$Tg <- T_g[as.character(r_g_theo$g)]

# 1. get results for the actual data - you may want to stop there if you are not interested in testing permutations
z_real <- z
#sanity_check_structs(z_real, morpho, partitions_sel, dist) #perform sanity checks, TO DO properly
if (map_type == "areas") {
  z_real <- select_points_in_partitions_areas(z_real, morpho, partitions_sel)
}
if (map_type == "sites") {
  z_real <- select_points_in_partitions_sites(z_real, partitions_sel)
}
# check distribution of categories across partitions
table(z_real$marks$marks.partition, z_real$marks$marks.cat)

countpg <- count_coocurrences(z_real, dist)
countpg$ct_pairs <- NULL
names(countpg)[names(countpg) == 'n'] <- 'real.counts'
r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))


# 2. add the random permutations
for (iter in 1:iter_n) {
  print(iter)
  z_shuf <- z
  z_shuf$marks$cat <- sample(z_shuf$marks$cat)
  # check where buildings fall
  #table(z_shuf$marks$marks.partition)/z_shuf$n
  
  #sanity_check_structs(z_shuf, morpho, partitions_sel, dist)
  if (map_type == "areas") {
    z_shuf <- select_points_in_partitions_areas(z_shuf, morpho, partitions_sel)
  }
  if (map_type == "sites") {
    z_shuf <- select_points_in_partitions_sites(z_shuf, partitions_sel)
  }
  countpg <- count_coocurrences(z_shuf, dist)
  countpg$ct_pairs <- NULL
  names(countpg)[names(countpg) == 'n'] <- paste('rand', iter, sep="_")
  r_g_theo <- left_join(r_g_theo, countpg, by = c('ord_ct1' = 'ord_ct1', 'ord_ct2' = 'ord_ct2', 'g' = 'g'))

}

#3. Save raw results from real data and permutation
perm <- paste("coocs",iter_n, "permutations.csv", sep="_")
write.csv(r_g_theo, paste(workdir, perm, sep="/"))
```

Checkpoint after iterations

```{r}
# Uncomment this line if you have already the time-consuming step if iterations of random permutations
#r_g_theo <- read.csv( paste(workdir, perm, sep="/")) 

#4. Convert to summary results
reps <- c("real.counts", paste(sep="", "rand_", 1:iter_n)) 
cols = c("ct_pairs", "ord_ct1", "ord_ct2", "g", "Tg")

df <- data.frame(matrix(ncol = 2, nrow = 0))
colnames(df) <- c('Hr_rel', 'rep')

for (rep_sim in reps) {
  print(rep_sim)
  r_allg <- r_g_theo[,c(cols, rep_sim)]
  ncoocs_by_ctpairs <- r_g_theo[,c(cols, rep_sim)] %>% # Count the total number of buildings from each co-occurence (across all partitions)
      replace(is.na(.), 0) %>% 
      group_by(ct_pairs) %>% summarise(sum_n= sum(get(rep_sim)))
  r_allg <- inner_join(r_allg, ncoocs_by_ctpairs, by="ct_pairs")
  r_allg <- arrange(r_allg, ct_pairs, g)
  r_all_g_fltr <-  prep_b4_calc_H(r_allg, rep_sim, min_coocs)
  dec_H_rep <- counts_to_H(r_all_g_fltr, T_all)
  r_all_g_fltr = dec_H_rep[[1]]
  dec_H_r = dec_H_rep[[2]]
  #store results from this rep in dataframe
  df <- rbind(df, data.frame(Hr_rel = dec_H_r$Hr_rel, rep = rep_sim))
  if (rep_sim == "real.counts"){ 
    df.real.counts <- dec_H_rep
  }
}

real.counts.ordered <- arrange(df.real.counts[[2]],Hr_rel)
write.csv(real.counts.ordered,  paste(workdir, "real.counts.Hr_rel.csv", sep="/"))
```

Decomposed RCE's
Sort by lowest (most signal) to highest (lowest signal=random) and display just the top pairs
```{r}
T_star <- T_all[[2]]
T_tot <- T_all[[3]]

realcnts <- df.real.counts[[1]] %>% 
  group_by(ct_pairs) %>% 
  summarise(sum_n = sum(n), RCE = sum(plogtg_p), 
            g1 = ifelse(g == 1, n, 0), 
            g2 = ifelse(g == 2, n, 0), 
            g3 = ifelse(g == 3, n, 0),
            g4 = ifelse(g == 4, n, 0),
            g5 = ifelse(g == 5, n, 0),
            g6 = ifelse(g == 6, n, 0),) %>%  
  arrange(,RCE)
realcnts <- realcnts %>%
  group_by(ct_pairs) %>%
  summarise(RCE = max(RCE), relRCE= (max(RCE) - log(T_star)) / (log(T_tot) - log(T_star)),
            n = max(sum_n),  
            g1 = sum(g1), 
            g2 = sum(g2), 
            g3 = sum(g3),
            g4 = sum(g4),
            g5 = sum(g5),
            g6 = sum(g6)
            ) %>% 
  arrange(,RCE)
#Fig. 2d
print(n=150, realcnts)
```

Summary statistics of permutations
Plotting of permutations against real data averaged per iteration (fig X3) and with each decomposed RCE (fig X4)
Decomposed RCE's of pairs (fig X5)

```{r}
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

df_by_rep <- df %>% group_by(rep) %>% summarise(mean_Hr_rel= mean(Hr_rel))
df_by_rep$test <- with(df_by_rep, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
permutations <- df_by_rep[df_by_rep$test == "permutation test",]

summary(permutations$mean_Hr_rel)
var(permutations$mean_Hr_rel)

#Fig 2b
pdf(paste(workdir, "fig2b.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54,
    bg = "white",
    colormodel = "cmyk",
    paper = "A4")
p <- ggplot(df_by_rep, aes(x=test, y=mean_Hr_rel)) + 
  geom_boxplot(size = 0.5, colour="black", fill="white", linewidth = 0.5, outlier.size = 0.1) +
  geom_jitter(size=0.1, alpha = 0.3) +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\nrooftypes","permuted\nrooftypes")) +
  scale_y_continuous(name =  "Relative RCE", breaks=seq(0.88,0.99,0.02), limits=c(0.885, 0.999), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 


#Fig. 2d
print(n=length(real.counts.ordered$ct_pairs), real.counts.ordered)

#Fig. 2c
min(df[df$test == "permutation test",]$Hr_rel)
df$test <- with(df, ifelse(rep == "real.counts" , "real.counts", "permutation test"))
pdf(paste(workdir, "fig2c.pdf", sep = "/"),
    width = 4 / 2.54 , height = 5 / 2.54,
    bg = "white",
    colormodel = "cmyk", 
    paper = "A4") 
p <- ggplot(df, aes(x=test, y=Hr_rel, fill=test)) +   geom_violin() +
  geom_point(size = 0.5, fill="black", alpha=1) + 
  geom_sina(alpha=0.01) +
  theme(legend.position="none") +
  xlab(element_blank()) + 
  scale_x_discrete(limits=c("real.counts", "permutation test"), labels=c("real\nrooftypes","permuted\nrooftypes")) +
  scale_y_continuous(name =  "Decomposed relative RCE", breaks=seq(0.6,0.99,0.1), limits=c(0.5, 1), labels = percent) +
  theme(axis.text.x = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.text.y = element_text(family = "Helvetica", size = 5, color = "black"),
        axis.title.x = element_text(family = "Helvetica", size = 7, color = "black"),
        axis.title.y = element_text(family = "Helvetica", size = 7, color = "black"), 
        panel.grid.minor = element_blank())
p
dev.off() 



```


